---
title: "Question 2: Computer arithmetics"
output:
  pdf_document: 
    keep_tex: true
editor_options: 
  markdown: 
    wrap: 72
tables: true
header-includes:
 \usepackage{float}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

```{r include=FALSE}

#' Custom Variance Calculation 
#'
#' This function calculates the variance of a numeric vector using a scaled formula to mitigate numerical instability. 
#' The data is first normalized by dividing by the maximum value, then the variance is computed using the standard formula, 
#' and finally scaled back to the original units.
#' 
#' @param x A numeric vector for which the sample variance is to be calculated.
#'
#' @return A numeric value representing the estimated variance of the input vector. 
myvar <- function(x){
  
  max_x <- max(x)
  x <- x / max_x
  
  
  n <- length(x)
  squaresum <- sum(x^2)
  sum_x <- sum(x)
  res <- (squaresum - (sum_x^2)/n)/(n-1)
  
  
  return(res * (max_x)^2)
}
```

```{r include=FALSE}
#' Stable Variance Calculation
#'
#' This function computes the sample variance of a numeric vector {x} in a numerically stable way, 
#' by centering each element around the mean before summing the squared differences.
#' 
#' @param x A numeric vector for which the sample variance is to be calculated.
#'
#' @returns A numeric value representing the estimated variance of the input vector. 
myvar_stable <- function(x) {
  n <- length(x)
  mean_x <- mean(x)
  sum_sq_diff <- sum((x - mean_x)^2)
  return(sum_sq_diff / (n - 1))
}
```

In `myvar()` function

$$
\mathrm{Var}(\vec{x}) = \frac{1}{n-1} \left( \sum_{i=1}^n x_i^2 - \frac{1}{n} \left( \sum_{i=1}^n x_i \right)^2 \right)
$$




```{r include=FALSE}
set.seed(12345)
x <- rnorm(10000,  10^8, 1)
```

Let $x$ be a vector containing 10,000 random numbers with a mean of
$10^8$ and a variance of 1. The first 10 elements of $x$ are shown
below:

```{r echo=FALSE}
print(head(x,10))
```

We define $Y$ as the difference between $\mathrm{myvar}(X_i)$ and
$\mathrm{var}(X_i)$, where $X_i = \{x_1, \dots, x_i\}$ for
$i = 1,\dots,10000$, and $\mathrm{var}(X_i)$ is the standard variance
function in R. Next, we plot how $Y_i$ changes as $i$ increases. The
resulting figure is shown below.

```{r echo=FALSE}
myvar_res <- numeric(10000)
var_res <- numeric(10000)
Y <- numeric(10000)
i_values <- 1:10000

for (i in i_values) {
  sub_x <- x[1:i]
  myvar_res[i] <- myvar(sub_x)
  var_res[i] <- var(sub_x)
  Y[i] <- myvar_res[i] - var_res[i]
  
}
```


```{r echo=FALSE, fig.height=4, fig.width=6}
plot(i_values,Y, main = "Difference between myvar(Xi) and var(Xi)",
     xlab = "i (Subset)", ylab = "Difference",lwd = 1,
      cex.lab = 1.2, cex.axis = 1.1, las = 1)

```
\clearpage
**Conclusion of plot**

The figure is a scatter plot with the subset size $i$ on the x-axis
(ranging from 1 to 10,000) and the difference $Y$ on the y-axis. Each point represents the difference
between the custom `myvar()` function and R’s built-in `var()` function
when using the first $i$ observations of the data. Key visual features
include:

1.  Multiple curved “bands”: The points cluster into several arcs, some
    in the positive region (up to around +3), and some in the negative
    region (down to around -3).

2.  A near-horizontal “line” or “band”: In some ranges of $i$, the
    points lie roughly on a constant difference level, not changing much
    as $i$ increases.

**Explanation of the Patterns**

The core reason is that the custom function `myvar()` first scales the
current subset by its maximum value (dividing by $max(x)$) before computing
the variance, and then rescales by multiplying back at the end. By
contrast, R’s `var()` function computes the variance directly on the
original data.

Because the data are on the order of $10^8$, directly computing
$\sum x^2 - \frac{(\sum x)^2}{n}$ can lead to catastrophic cancellation
(subtracting two large numbers). Scaling the data by the maximum value
before computing the variance can mitigate some numerical issues, but
also creates these distinct patterns:

1.  If new data points do not exceed the current maximum, the scaling
    factor in `myvar()` remains unchanged, so the difference is a
    constant amount. On the plot, this appears as a near-horizontal line
    or “band.”

2.  As soon as a larger data point arrives, the maximum value jumps,
    which changes the scaling factor in `myvar()`. This causes a sudden
    shift in the difference. Subsequently, for a range of $i$ values
    with that fixed new maximum, the difference will evolve smoothly
    again, forming a curved arc, until the next time a bigger maximum is
    found.

To further refine our variance estimator, we will adopt a formula that
exactly matches R’s built-in `var()`. Below, we present the
revised approach and demonstrate how it produces identical results to
`var()`.

The improved formula is:

$$
\mathrm{Var}(x) = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2 
$$

Still, we define $Y$ as the difference between  $\mathrm{myvar\_stable}(X_i)$ and
$\mathrm{var}(X_i)$, where $X_i = \{x_1, \dots, x_i\}$ for
$i = 1,\dots,10000$, and $\mathrm{var}(X_i)$ is the standard variance
function in R. Next, we plot how $Y_i$ changes as $i$ increases. The
resulting figure is shown below.

```{r echo=FALSE}
myvar_stable_res <- numeric(10000)
var_res_2 <- numeric(10000)
Y_stable <- numeric(10000)
for (i in i_values) {
  sub_x <- x[1:i]
  myvar_stable_res[i] <- myvar_stable(sub_x)
  var_res_2[i] <- var(sub_x)
  Y_stable[i] <- myvar_stable_res[i] - var_res_2[i]
  
}
```


```{r echo=FALSE, fig.height=4, fig.width=6}
plot(i_values,Y_stable, main = "Difference between myvar_stable(Xi) and var(Xi)",
     xlab = "i (Subset)", ylab = "Difference",ylim = c(-0.01,0.01),lwd = 1,
      cex.lab = 1.2, cex.axis = 1.1
     )
```
  
The improved formula is numerically more stable because it first “centers” each observation $x_i$ by subtracting the mean $\bar{x}$ before squaring. This step avoids the large intermediate values that can result when squaring data with very large means, reducing potential floating-point round‐off errors. As a result, the variance estimate remains accurate even for data with large magnitudes.


## Appendix

```r

#' Custom Variance Calculation 
#'
#' This function calculates the variance of a numeric vector using a scaled formula to mitigate numerical instability. 
#' The data is first normalized by dividing by the maximum value, then the variance is computed using the standard formula, 
#' and finally scaled back to the original units.
#' 
#' @param x A numeric vector for which the sample variance is to be calculated.
#'
#' @return A numeric value representing the estimated variance of the input vector. 
myvar <- function(x){
  
  max_x <- max(x)
  x <- x / max_x
  
  
  n <- length(x)
  squaresum <- sum(x^2)
  sum_x <- sum(x)
  res <- (squaresum - (sum_x^2)/n)/(n-1)
  
  
  return(res * (max_x)^2)
}


#' Stable Variance Calculation
#'
#' This function computes the sample variance of a numeric vector {x} in a numerically stable way, 
#' by centering each element around the mean before summing the squared differences.
#' 
#' @param x A numeric vector for which the sample variance is to be calculated.
#'
#' @returns A numeric value representing the estimated variance of the input vector. 
myvar_stable <- function(x) {
  n <- length(x)
  mean_x <- mean(x)
  sum_sq_diff <- sum((x - mean_x)^2)
  return(sum_sq_diff / (n - 1))
}


myvar_res <- numeric(10000)
var_res <- numeric(10000)
Y <- numeric(10000)
i_values <- 1:10000

for (i in i_values) {
  sub_x <- x[1:i]
  myvar_res[i] <- myvar(sub_x)
  var_res[i] <- var(sub_x)
  Y[i] <- myvar_res[i] - var_res[i]
  
}


myvar_stable_res <- numeric(10000)
var_res_2 <- numeric(10000)
Y_stable <- numeric(10000)
for (i in i_values) {
  sub_x <- x[1:i]
  myvar_stable_res[i] <- myvar_stable(sub_x)
  var_res_2[i] <- var(sub_x)
  Y_stable[i] <- myvar_stable_res[i] - var_res_2[i]
  
}
```
