---
title: "Question 1: Maximization of a function in one variable"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
tables: true
header-includes:
 \usepackage{float}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```


```{r include=FALSE}
# -------- OPTIMIZATION ALGORITHMS --------


#' Bisection Method for Finding a Maximum
#'
#' This function uses the bisection method to find the maximum value of a given function `g` 
#' within the interval `[a, b]`.
#'
#' @param g A function representing the objective function to maximize.
#' @param g.prime A function representing the derivative of `g`.
#' @param a A numeric value specifying the lower bound of the interval.
#' @param b A numeric value specifying the upper bound of the interval.
#' @param tol A numeric value specifying the tolerance for stopping the iteration. Default is 0.001.
#' 
#' @return A data frame containing the iteration history with columns for the iteration number, 
#' the current `x` value, `g(x)`, and `g.prime(x)`.
mybisection <- function(g, g.prime, a, b, tol = 0.001) {
  
  xt <- c(0,0)
  xt[1] <- (a+b)/2
  
  history_df <- data.frame(iter = numeric(0), x = numeric(0), g = numeric(0), g.prime = numeric(0))
  
  i <- 1
  while (abs(xt[2] - xt[1]) > tol) {
    
    
    deriv_a <- g.prime(a)
    deriv_b <- g.prime(b)
    deriv_xt <- g.prime(xt[1])
    
    if((deriv_a * deriv_xt) < 0){
      b <- xt[1]
      xt[2] <- xt[1]
      xt[1] <- (a+b)/2
    }
    else{
      a <- xt[1]
      xt[2] <- xt[1]
      xt[1] <- (a+b)/2
    }
    
    # Save current values for x, g and gprime.    
    history_df[nrow(history_df) + 1, ] <- c(i, xt[1], g(xt[1]), g.prime(xt[1]))
    i <- i + 1

  }
  return(history_df)
}


#' Secant Method for Finding a Root
#'
#' This function uses the secant method to find a root of the given function `g`, 
#' starting from an initial guess and iterating until the specified tolerance is reached.
#'
#' @param g A function representing the objective function for which the root is sought.
#' @param g.prime A function representing the derivative of `g`.
#' @param initial.guess A numeric value specifying the initial guess for the root.
#' @param tol A numeric value specifying the tolerance for stopping the iteration. Default is 0.001.
#' 
#' @return A data frame containing the iteration history with columns for the iteration number, 
#' the current `x` value, `g(x)`, and `g.prime(x)`.
mysecant <- function(g, g.prime, a, b, tol = 0.001) {
  
  xt_secant <- c(a, b)
  history_df <- data.frame(
    iter = numeric(0), 
    x = numeric(0), 
    g = numeric(0),
    g.prime = numeric(0)
    )
  
  i <- 1
  while (abs(xt_secant[2] - xt_secant[1]) > tol){
    

    deriv_xt <- g.prime(xt_secant[1])
    deriv_xt_minus1 <- g.prime(xt_secant[2])
    x_tplus1 <- xt_secant[1] - deriv_xt * (xt_secant[1]-xt_secant[2])/(deriv_xt-deriv_xt_minus1)
    xt_secant[2] <- xt_secant[1]
    xt_secant[1] <- x_tplus1
    
    # Save current values for x, g and gprime. 
    history_df[nrow(history_df) + 1, ] <- c(i, xt_secant[1], g(xt_secant[1]), g.prime(xt_secant[1]))
    i <- i + 1
    
  }
  return(history_df)
  
  
}
```


Consider the function

$$
g(x) = \frac{\log{(x+1)}}{x^{\frac{3}{2}} + 1}
$$

A plot of this function in the $[0, 4]$ interval is shown in the following image.


```{r, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, out.extra='trim=0 0 0 0, clip'}
# Adjust margins to reduce white space
par(mar = c(4, 4, 2, 1))  # Bottom, left, top, right margins (in lines)

g <- function(x) {
  num = log(x + 1)
  den = x^(3/2) + 1
  return(num / den)
}

xvalues <- seq(0, 4, length.out = 50)
gvalues <- g(xvalues)

# Define custom y-axis ticks
ytick_interval <- 0.1  # Set desired interval between ticks
yticks <- seq(0, max(gvalues), by = ytick_interval)  # Create custom ticks

# Create the plot
plot(
  xvalues, gvalues, 
  type = "l", col = "blue", lwd = 2,
  xlab = "x", ylab = "g(x)", 
  cex.lab = 1.2, cex.axis = 1.1, las = 1,
  yaxt = "n"  # Suppress default y-axis
)

# Add custom y-axis ticks
axis(2, at = yticks, labels = round(yticks, 2), las = 1)  # Use rounded labels
grid(col = "gray", lty = "dotted", lwd = 1)

# Reset margins after the plot if needed for other plots
par(mar = c(5, 4, 4, 2) + 0.1)

```

Its derivative, $g'(x)$, is given by

$$
g'(x) = \frac{1}{\left(x + 1\right) \left(x^{\frac{3}{2}} + 1\right)} - \frac{3 \sqrt{x} \ln\left(x + 1\right)}{2 \left(x^{\frac{3}{2}} + 1\right)^{2}}
$$
whose plot in the interval $[0, 4]$ is shown below.

```{r, fig.height=2.5, fig.width=4, fig.align="center", echo=FALSE, out.extra='trim=0 0 0 0, clip'}

par(mar = c(4, 4, 2, 1))  # Bottom, left, top, right margins (in lines)


gprime <- function(x) {
  return(1 / ((x + 1) * (x^(3 / 2) + 1)) - (3 * sqrt(x) * log(x + 1)) / (2 * (x^(3 / 2) + 1)^2))
}

xvalues <- seq(0, 4, length.out = 50)
gprimevalues <- gprime(xvalues)

# Define custom y-axis ticks
ytick_interval <- 0.2  # Set desired interval between ticks
yticks <- seq(0, max(gprimevalues), by = ytick_interval)  # Create custom ticks


plot(
  xvalues, gprimevalues, 
  type = "l", col = "blue", lwd = 2,
  xlab = "x", ylab = "g'(x)", 
  cex.lab = 1.2, cex.axis = 1.1, las = 1,
  yaxt = "n"  # Suppress default y-axis
)

# Add custom y-axis ticks
axis(2, at = yticks, labels = round(yticks, 2), las = 1)  # Use rounded labels
grid(col = "gray", lty = "dotted", lwd = 1)

# Reset margins after the plot if needed for other plots
par(mar = c(5, 4, 4, 2) + 0.1)


abline(h = 0, col = "black", lty = "dashed", lwd = 2)
```

Notice a black dashed horizontal line at $g'(x) = 0$ is included in the plot
to visually inspect where in the x-axis the derivative becomes $0$.

## Bisection method

Next, we present the evolution of the **bisection** method (implementation given in 
the **Appendix** section) for finding the value of $x$ that maximizes $g(x)$.


```{r echo=FALSE}
history_df <- mybisection(g, gprime, 0, 2)
history_df <- round(history_df, 3)
knitr::kable(
  history_df,
  caption = "Evolution of the bisection algorithm for $a=0$ and $b=2$."
)

```

The bisection algorithm was run using $a=0$ and $b=2$, where $a$ and $b$ denote
the lower and upper bounds, respectively. The next table shows the results for 
different values of $a$ and $b$, whether they converge to the true maximizer
or not and the number of iterations until the threshold is met.



```{r include=FALSE}
avalues <- seq(0, 1.4, by = .2)
bvalues <- avalues + 2
true_minimizer <- 0.962

results <- data.frame(
  niter = numeric(0), 
  a = numeric(0), 
  b = numeric(0), 
  x = numeric(0),
  success = logical(0)
)

for (i in 1:length(avalues)) {
  
  a <- avalues[i]
  b <- bvalues[i]
  
  history <- mybisection(g, gprime, a, b, 1e-5)
  
  niter <- nrow(history)
  lastrow <- tail(history, 1)
  minimizer <- lastrow$x
  
  success <- abs(minimizer - true_minimizer) < 1e-3
  
  
  results[nrow(results) + 1, ] <- c(niter, a, b, minimizer, success)
}
```


```{r echo=FALSE}
results <- round(results, 3)
knitr::kable(
  results,
  caption = "Bisection algorithm for different starting values $a$ and $b$."
)
```

Notice the optimizer is sensitive to starting value conditions in the follwing
way: as long as the starting interval $[a, b]$ contains the true maximizer 
$x^* \approx 0.961$, the algorithm converges correctly. Otherwise, it returns
the endpoint $b$ since a change of polarity is never found.


## Secant method

Now, we turn our attention to the **secant** method (implementation given in 
the **Appendix** section). The following table contains the result when
$a = 0$ and $b = 0$ for the same function $g(x)$.


```{r echo=FALSE}
history_df <- mysecant(g, gprime, 0, 1)
history_df <- round(history_df, 3)
knitr::kable(
  history_df,
  caption = "Evolution of the secant algorithm for $x_0=0$"
)

```

Similarly than before, we inspect how the algorithm behaves for different
starting values.

```{r include=FALSE}
avalues <- seq(0, 2, by = .2)
bvalues <- avalues + 1
true_minimizer <- 0.962

results <- data.frame(
  niter = numeric(0), 
  a = numeric(0), 
  b = numeric(0), 
  x = numeric(0),
  success = logical(0)
)

for (i in 1:length(avalues)) {
  
  a <- avalues[i]
  b <- bvalues[i]
  
  newrow <- tryCatch(
    {
      # Try to run function `mysecant`.
      history <- mysecant(g, gprime, a, b, 1e-3)
      niter <- nrow(history)
      lastrow <- tail(history, 1)
      minimizer <- lastrow$x
      success <- abs(minimizer - true_minimizer) < 1e-3
      
      c(niter, a, b, minimizer, success)
    },
    error = function(cond) {
      c(NA, a, b, NA, FALSE)
    }
    
  )
  
  
  results[nrow(results) + 1, ] <- newrow
}

```

```{r echo=FALSE}
knitr::kable(
  round(results, 3),
  caption = "Secant algorithm for different starting values."
)
```

Since $g'(x)$ is nearly flat for values greater than, say, $2$, the secant 
method can become unstable when using starting values in this region. This 
happens because the secant lines connecting successive iterates will have 
small slopes, meaning the denominator in the secant update formula will be 
small. As a result, the step size becomes large, potentially causing the 
iterates to diverge. The direction of divergence depends on whether the small 
slope is positive or negative, leading to large positive or negative values 
of $x$ in subsequent iterations. NAN values are produces when the direction
of divergence is negative since the expression for $g'(x)$ includes
functions not defined for negative values.


## Which one should be used?

At iteration $t$, the small negative slope in $g'(x)$ caused $x_{t+1}$ to become
negative. This led to an unfeasible evaluation since the function involves
terms like $\sqrt{x}$ and $\ln(x)$ which are undefined for negative numbers. 
In this case, the bisection method will be a better choice because its search space
is always bounded by the initial range $[a, b]$. 

This reasoning leads to a way to decide between both methods in a more general 
setting: if the objective function is not defined for some region 
(e.g., $x < 0$), then it would be better to constraint the search space and prefer the bisection method. On the other hand, if the function is well-defined over the entire domain, the faster convergence of the secant method makes it a more efficient choice.


## Appendix

```r
# -------- FUNCTION DEFINITION --------

g <- function(x) {
  num = log(x + 1)
  den = x^(3/2) + 1
  return(num / den)
}

gprime <- function(x) {
  return(1 / ((x + 1) * (x^(3 / 2) + 1)) - (3 * sqrt(x) * log(x + 1)) / (2 * (x^(3 / 2) + 1)^2))
}

# -------- OPTIMIZATION ALGORITHMS --------


#' Bisection Method for Finding a Maximum
#'
#' This function uses the bisection method to find the maximum value of a given function `g` 
#' within the interval `[a, b]`.
#'
#' @param g A function representing the objective function to maximize.
#' @param g.prime A function representing the derivative of `g`.
#' @param a A numeric value specifying the lower bound of the interval.
#' @param b A numeric value specifying the upper bound of the interval.
#' @param tol A numeric value specifying the tolerance for stopping the iteration. Default is 0.001.
#' 
#' @return A data frame containing the iteration history with columns for the iteration number, 
#' the current `x` value, `g(x)`, and `g.prime(x)`.
mybisection <- function(g, g.prime, a, b, tol = 0.001) {
  
  xt <- c(0,0)
  xt[1] <- (a+b)/2
  
  history_df <- data.frame(iter = numeric(0), x = numeric(0), g = numeric(0), g.prime = numeric(0))
  
  i <- 1
  while (abs(xt[2] - xt[1]) > tol) {
    
    
    deriv_a <- g.prime(a)
    deriv_b <- g.prime(b)
    deriv_xt <- g.prime(xt[1])
    
    if((deriv_a * deriv_xt) < 0){
      b <- xt[1]
      xt[2] <- xt[1]
      xt[1] <- (a+b)/2
    }
    else{
      a <- xt[1]
      xt[2] <- xt[1]
      xt[1] <- (a+b)/2
    }
    
    # Save current values for x, g and gprime.    
    history_df[nrow(history_df) + 1, ] <- c(i, xt[1], g(xt[1]), g.prime(xt[1]))
    i <- i + 1

  }
  return(history_df)
}


#' Secant Method for Finding a Root
#'
#' This function uses the secant method to find a root of the given function `g`, 
#' starting from an initial guess and iterating until the specified tolerance is reached.
#'
#' @param g A function representing the objective function for which the root is sought.
#' @param g.prime A function representing the derivative of `g`.
#' @param initial.guess A numeric value specifying the initial guess for the root.
#' @param tol A numeric value specifying the tolerance for stopping the iteration. Default is 0.001.
#' 
#' @return A data frame containing the iteration history with columns for the iteration number, 
#' the current `x` value, `g(x)`, and `g.prime(x)`.
mysecant <- function(g, g.prime, a, b, tol = 0.001) {
  
  xt_secant <- c(a, b)
  history_df <- data.frame(
    iter = numeric(0), 
    x = numeric(0), 
    g = numeric(0),
    g.prime = numeric(0)
    )
  
  i <- 1
  while (abs(xt_secant[2] - xt_secant[1]) > tol){
    

    deriv_xt <- g.prime(xt_secant[1])
    deriv_xt_minus1 <- g.prime(xt_secant[2])
    x_tplus1 <- xt_secant[1] - deriv_xt * (xt_secant[1]-xt_secant[2])/(deriv_xt-deriv_xt_minus1)
    xt_secant[2] <- xt_secant[1]
    xt_secant[1] <- x_tplus1
    
    # Save current values for x, g and gprime. 
    history_df[nrow(history_df) + 1, ] <- c(i, xt_secant[1], g(xt_secant[1]), g.prime(xt_secant[1]))
    i <- i + 1
    
  }
  return(history_df)
}



# --------- BISECTION FOR DIFFERENT STARTING VALUES ---------

avalues <- seq(0, 1.4, by = .2)
bvalues <- avalues + 2
true_minimizer <- 0.962

results <- data.frame(
  niter = numeric(0), 
  a = numeric(0), 
  b = numeric(0), 
  x = numeric(0),
  success = logical(0)
)

for (i in 1:length(avalues)) {
  
  a <- avalues[i]
  b <- bvalues[i]
  
  history <- mybisection(g, gprime, a, b, 1e-5)
  
  niter <- nrow(history)
  lastrow <- tail(history, 1)
  minimizer <- lastrow$x
  
  success <- abs(minimizer - true_minimizer) < 1e-3
  
  
  results[nrow(results) + 1, ] <- c(niter, a, b, minimizer, success)
}


# --------- SECANT FOR DIFFERENT STARTING VALUES ---------

avalues <- seq(0, 2, by = .2)
bvalues <- avalues + 1
true_minimizer <- 0.962

results <- data.frame(
  niter = numeric(0), 
  a = numeric(0), 
  b = numeric(0), 
  x = numeric(0),
  success = logical(0)
)

for (i in 1:length(avalues)) {
  
  a <- avalues[i]
  b <- bvalues[i]
  
  newrow <- tryCatch(
    {
      # Try to run function `mysecant`.
      history <- mysecant(g, gprime, a, b, 1e-3)
      niter <- nrow(history)
      lastrow <- tail(history, 1)
      minimizer <- lastrow$x
      success <- abs(minimizer - true_minimizer) < 1e-3
      
      c(niter, a, b, minimizer, success)
    },
    error = function(cond) {
      c(NA, a, b, NA, FALSE)
    }
    
  )
  
  
  results[nrow(results) + 1, ] <- newrow
}

```


