---
title: "Question 2: Gibbs sampling"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
tables: true
header-includes:
 \usepackage{float}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

```{r include=FALSE}
#' Gibbs sampling for bivariate distribution f(x1,x2)
#'
#' @param w Parameter of the density function
#' @param n Number of samples
#'
#' @returns A matrix which contains the samples
gibbs_sampling <- function(w,n) {
  x1 <- 0
  x2 <- 0
  samples <- matrix(0, nrow = n, ncol = 2)
  
  for (i in 1:n) {

    x1_min <- -(w/2) * x2 - sqrt(1-(1-w^2/4)*x2^2)
    x1_max <- -(w/2) * x2 + sqrt(1-(1-w^2/4)*x2^2)
    x1 <- runif(1, x1_min, x1_max)
    
    x2_min <- -(w/2) * x1 - sqrt(1-(1-w^2/4)*x1^2)
    x2_max <- -(w/2) * x1 + sqrt(1-(1-w^2/4)*x1^2)
    x2 <- runif(1, x2_min, x2_max)
    
    samples[i, ] <- c(x1, x2)
  }
  
  return(samples)
}


#' Gibbs sampling for bivariate distribution f(u1,u2)
#'
#' @param w Parameter of the density function
#' @param n Number of samples
#'
#' @returns A matrix which contains the samples
gibbs_sampling_U <- function(w, n) {
  U1 <- 0
  U2 <- 0
  samples_U <- matrix(0, nrow = n, ncol = 2)
  
  for (i in 1:n) {
    
    molecular1 <- 4 - (2 + w)*U2^2
    rangeU1 <- sqrt(molecular1 / (2 - w))
    U1_min <- -rangeU1
    U1_max <-  rangeU1
    U1 <- runif(1, U1_min, U1_max)
    
    molecular2 <- 4 - (2 - w)*U1^2
    rangeU2 <- sqrt(molecular2 / (2 + w))
    U2_min <- -rangeU2
    U2_max <-  rangeU2
    U2 <- runif(1, U2_min, U2_max)
    
    samples_U[i, ] <- c(U1, U2)
  }
  

  return(samples_U)
}
```

For bivariate distribution with density 
$$
f(x_1, x_2) \propto \mathbb{1} \{ x_1^2 + w x_1 x_2 + x_2^2 < 1 \}
$$
The boundaries of the region is shown below:

```{r fig.height=4, fig.width=5, fig.align="center",echo=FALSE}
w  <- 1.999
xv <- seq(-1, 1, by=0.01) * 1/sqrt(1-w^2/4)  
plot(xv, xv, type="n", xlab=expression(x[1]), ylab=expression(x[2]), las=1)
# ellipse
lines(xv, -(w/2)*xv-sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
lines(xv, -(w/2)*xv+sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
```

With the formula we can calculate the conditional distribution of $X_1$ given $X_2$ is uniform on:

$$
\left[ -\frac{w}{2} x_2 - \sqrt{1 - \left(1 - \frac{w^2}{4}\right) x_2^2}, 
       -\frac{w}{2} x_2 + \sqrt{1 - \left(1 - \frac{w^2}{4}\right) x_2^2} \right ]
$$

and the conditional distribution of $X_2$ given $X_1$ is uniform on:

$$
\left[ -\frac{w}{2} x_1 - \sqrt{1 - \left(1 - \frac{w^2}{4}\right) x_1^2}, 
       -\frac{w}{2} x_1 + \sqrt{1 - \left(1 - \frac{w^2}{4}\right) x_1^2} \right ]
$$
After generating 1000 random vectors with Gibbs sampling, the plot is shown below:

```{r fig.height=4, fig.width=5, fig.align="center",echo=FALSE}
set.seed(12345)
samples <- gibbs_sampling(1.999,1000)
w  <- 1.999
xv <- seq(-1, 1, by=0.01) * 1/sqrt(1-w^2/4)  
plot(xv, xv, type="n", xlab=expression(x[1]), ylab=expression(x[2]), las=1)
# ellipse
lines(xv, -(w/2)*xv-sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
lines(xv, -(w/2)*xv+sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
points(samples[,1], samples[,2], pch = 20, cex = 0.5, col = "blue")
```
Based on the sample, the result $P(X_1 > 0)$ is:

```{r echo=FALSE}
mean(samples[,1] > 0)
```
But repeat this a few times, we can get the true result is about:
```{r echo=FALSE}
res <- numeric(50)

for (i in 1:50) {
  sample <- gibbs_sampling(1.999,1000)
  res[i] <- mean(sample[,1] > 0)
}

mean(res)
```
As $w$ is close to 2, the ellipse becomes more and more slender.  This makes the Gibbs chain mix more slowly. The conditional ranges in one direction gets narrow, so the sampler makes very small moves in each step. This means it takes many iterations to explore the whole ellipse.

We might transform the variable $X$ and generate $U = (U_1,U_2)) = (X_1 -X_2,X_1+X_2)$ instead. In this case, the boundaries of this region is:

```{r fig.height=4, fig.width=5, fig.align="center",echo=FALSE}
u1v <- seq(-1, 1, by = 0.01) * sqrt(4 / (2 - w))
u2_plus  <-  sqrt( pmax(0, 4 - (2 - w)*u1v^2 ) / (2 + w) )
u2_minus <- -u2_plus
plot(u1v, u2_plus, type="n", xlab=expression(u[1]), ylab=expression(u[2]), 
     asp = 1, las = 1, 
     )
lines(u1v,  u2_plus,  lwd=2, col=8)
lines(u1v,  u2_minus, lwd=2, col=8)
```

Then we generate 1000 random vectors with Gibbs sampling for $U$ and the plot is:

```{r fig.height=4, fig.width=5, fig.align="center",echo=FALSE}
samples_U <- gibbs_sampling_U(1.999,1000)
u1v <- seq(-1, 1, by = 0.01) * sqrt(4 / (2 - w))
u2_plus  <-  sqrt( pmax(0, 4 - (2 - w)*u1v^2 ) / (2 + w) )
u2_minus <- -u2_plus
plot(u1v, u2_plus, type="n", xlab=expression(u[1]), ylab=expression(u[2]), 
     asp = 1, las = 1, 
    )
lines(u1v,  u2_plus,  lwd=2, col=8)
lines(u1v,  u2_minus, lwd=2, col=8)

points(samples_U[,1], samples_U[,2], pch = 20, cex = 0.5, col = "red")
```
Next, we use the transformation to get samples $X$ and plot them:

```{r fig.height=4, fig.width=5, fig.align="center",echo=FALSE}
samples_X <- cbind( (samples_U[,1] + samples_U[,2]) / 2,
                    (samples_U[,2] - samples_U[,1]) / 2 )
w  <- 1.999
xv <- seq(-1, 1, by=0.01) * 1/sqrt(1-w^2/4)  
plot(xv, xv, type="n", xlab=expression(x[1]), ylab=expression(x[2]), las=1)
# ellipse
lines(xv, -(w/2)*xv-sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
lines(xv, -(w/2)*xv+sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
points(samples_X[,1], samples_X[,2], pch = 20, cex = 0.5, col = "blue")
```
Based on this sample, $P(X_1 > 0)$ is:
```{r echo=FALSE}
mean(samples_X[,1] > 0)
```
After repeating a few times, $P(X_1 > 0)$ is:
```{r echo=FALSE}
res2 <- numeric(50)

for (i in 1:50) {
  sample <- gibbs_sampling_U(1.999,1000)
  samples_X <- cbind( (sample[,1] + sample[,2]) / 2,
                    (sample[,2] - sample[,1]) / 2 )
  res2[i] <- mean(samples_X[,1] > 0)
}

mean(res2)
```
Compare the result with Part c, Transformation to $(U_1,U_2)$ makes sampling more uniform. The result is that sample points are almost spread across the entire area.

A single sample, the result of $P(X_1 > 0)$ from Transformation $U$ is more stable, which cloese to 1/2. But since the elliptical area is symmetrical, the results after multiple samples will tend to be 1/2.


## Appendix
```{r echo=TRUE, eval=FALSE}
#' Gibbs sampling for bivariate distribution f(x1,x2)
#'
#' @param w Parameter of the density function
#' @param n Number of samples
#'
#' @returns A matrix which contains the samples
gibbs_sampling <- function(w,n) {
  x1 <- 0
  x2 <- 0
  samples <- matrix(0, nrow = n, ncol = 2)
  
  for (i in 1:n) {

    x1_min <- -(w/2) * x2 - sqrt(1-(1-w^2/4)*x2^2)
    x1_max <- -(w/2) * x2 + sqrt(1-(1-w^2/4)*x2^2)
    x1 <- runif(1, x1_min, x1_max)
    
    x2_min <- -(w/2) * x1 - sqrt(1-(1-w^2/4)*x1^2)
    x2_max <- -(w/2) * x1 + sqrt(1-(1-w^2/4)*x1^2)
    x2 <- runif(1, x2_min, x2_max)
    
    samples[i, ] <- c(x1, x2)
  }
  
  return(samples)
}


#' Gibbs sampling for bivariate distribution f(u1,u2)
#'
#' @param w Parameter of the density function
#' @param n Number of samples
#'
#' @returns A matrix which contains the samples
gibbs_sampling_U <- function(w, n) {
  U1 <- 0
  U2 <- 0
  samples_U <- matrix(0, nrow = n, ncol = 2)
  
  for (i in 1:n) {
    
    molecular1 <- 4 - (2 + w)*U2^2
    rangeU1 <- sqrt(molecular1 / (2 - w))
    U1_min <- -rangeU1
    U1_max <-  rangeU1
    U1 <- runif(1, U1_min, U1_max)
    
    molecular2 <- 4 - (2 - w)*U1^2
    rangeU2 <- sqrt(molecular2 / (2 + w))
    U2_min <- -rangeU2
    U2_max <-  rangeU2
    U2 <- runif(1, U2_min, U2_max)
    
    samples_U[i, ] <- c(U1, U2)
  }
  

  return(samples_U)
}

w  <- 1.999
xv <- seq(-1, 1, by=0.01) * 1/sqrt(1-w^2/4)  
plot(xv, xv, type="n", xlab=expression(x[1]), ylab=expression(x[2]), las=1)
# ellipse
lines(xv, -(w/2)*xv-sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
lines(xv, -(w/2)*xv+sqrt(1-(1-w^2/4)*xv^2), lwd=2, col=8)
set.seed(12345)
samples <- gibbs_sampling(1.999,1000)
mean(samples[,1] > 0)
points(samples[,1], samples[,2], pch = 20, cex = 0.5, col = "blue")

samples_U <- gibbs_sampling_U(1.999,1000)
samples_X <- cbind( (samples_U[,1] + samples_U[,2]) / 2,
                    (samples_U[,2] - samples_U[,1]) / 2 )


u1v <- seq(-1, 1, by = 0.01) * sqrt(4 / (2 - w))
u2_plus  <-  sqrt( pmax(0, 4 - (2 - w)*u1v^2 ) / (2 + w) )
u2_minus <- -u2_plus
plot(u1v, u2_plus, type="n", xlab=expression(u[1]), ylab=expression(u[2]), 
     asp = 1, las = 1, )
lines(u1v,  u2_plus,  lwd=2, col=8)
lines(u1v,  u2_minus, lwd=2, col=8)

points(samples_U[,1], samples_U[,2], pch = 20, cex = 0.5, col = "red")


```

