---
title: "Question 1: Computations with Metropolis–Hastings"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
tables: true
header-includes:
 \usepackage{float}
 \setlength{\textfloatsep}{0pt}

---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

```{r include=FALSE}
library(ggplot2)
library(gridExtra)
```


```{r include=FALSE}
densityTarget <- function(x) ifelse(x > 0, (1/120) * x^5 * exp(-x), 0)

#' Metropolis-Hastings Algorithm
#'
#' @param n integer.
#'   The number of samples to generate.
#' @param dtarget function.
#'   The target distribution's density function. 
#'   Must be a function with signature x -> dtarget(x).
#' @param dprop function.
#'   Density function for the proposed conditional distribution. 
#'   Must be a function with signature x, y -> dprop(x, y), where y is the 
#'   conditioning event and x is a value in the support.
#' @param rprop function.
#'   Random generator for the proposed conditional distribution. 
#'   Must be a function with signature x -> rprop(x), where x is the 
#'   conditioning event.
#' @param x0 numeric.
#'   Initial value for the sampling process. Must be in the support of `dprop`.
#'
#' @return numeric vector.
#'   A vector of `n` samples drawn from the target distribution.
rmh <- function(n, densityTarget, densityProposal, randomProposal, x0) {
  
  draws <- numeric(n)
  new_x <- x0
   
  for (i in 1:n) {
    
    last_x <- new_x
    
    # Sample a candidate from the proposal distribution.
    y <- randomProposal(last_x)
    
    # Compute the MH ratio.
    R <- (densityTarget(y) * densityProposal(last_x, y)) / (densityTarget(last_x) * densityProposal(y, last_x))
    
    # Decide new sample
    new_x <- ifelse(
      R >= 1, 
      y,
      ifelse(runif(1) < R, y, last_x)
      )
    
    draws[i] <- new_x
  }
  
  return(draws)
}
```



We are given the target density

$$
f(x) = \frac{1}{120} x^5 e^{-x} \quad x > 0
$$

In what follows ww will use the Metropolis-Hasting algorithm (MH) to generate
random samples from this density.



## Normal distribution

```{r include=FALSE}
normalDensityProposal <- function(x, y) dnorm(x, mean = y, sd = 0.1)
normalRandomProposal <- function(x) rnorm(n = 1, mean = x, sd = 0.1)
```

Here, we apply the MH algorithm using $\mathcal{N}(\mu = X_t, \sigma = 0.1)$ as the
proposed conditional density $g_{X_{t+1} | X_t}$. That is, 

$$
g_{X_{t+1} | X_t}(x_{t+1} | x_t) = \mathcal{N}(x_{t+1} | \mu = x_t, \sigma = 0.1)
$$

The following figure shows four runs of length $10000$ using different 
starting values.


```{r include=FALSE}
x1 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, 0.1)
x2 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, 0.2)
x3 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, 0.3)
x4 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, 0.4)
```

```{r echo=FALSE, fig.height=2, fig.width=6, fig.align="center", fig.cap = "Multiple execution of the MH algorithm using $\\mathcal{N}(\\mu = X_t, \\sigma = 0.1)$ as the proposal distribution."}
# Create a 2x2 grid layout
par(mfrow = c(2, 2), mar = c(3, 3, 0.5, 0.5), oma = c(0, 0, 0, 0))  # Adjust margins

# Plot the first vector (x1)
plot(x1, type = "l", main = "",xlab = "Iteration",  ylab = "Value", col = "black")

# Plot the second vector (x2)
plot(x2, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Plot the third vector (x3)
plot(x3, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Plot the fourth vector (x4)
plot(x4, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Reset layout to single plot (optional)
par(mfrow = c(1, 1))
```



In addition,the following plot overlaps the histogram from a single chain of
length $100,000$ with the true target density $f$.


```{r fig.height=2, fig.width=3, fig.align="center", fig.cap="Histogram of MH deviates values using $\\mathcal{N}(\\mu = X_t, \\sigma = 0.1)$ as the proposal distribution. The red curve shows the target (true) density.", echo=FALSE}
par(mar = c(3, 3, 1, 1))  # Adjust margins: c(bottom, left, top, right)
x <- rmh(100000, densityTarget, normalDensityProposal, normalRandomProposal, 0.1)
hist(x, probability = TRUE, breaks = 20, col = "skyblue", main = "")
curve(densityTarget(x), add = TRUE, col = "red", lwd = 2)
xNormal <- x
```


**What can you guess about the convergence of the chain? If there
is a burn–in period, what can be the size of this period? What is the acceptance rate?**


## Chi-square distribution

```{r include=FALSE}
chiDensityProposal <- function(x, y) dchisq(x, df = floor(y + 1))
chiRandomProposal <- function(x) rchisq(n = 1, df = floor(x + 1))
```


Now, we use $\chi(\lfloor{X_t + 1}\rfloor)$ as the proposal distribution where
$\lfloor{\cdot}\rfloor$ is the floor function. The results are shown in the 
following figures.

```{r include=FALSE}
x1 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, 0.1)
x2 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, 0.2)
x3 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, 0.3)
x4 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, 0.4)
```

```{r echo=FALSE, fig.height=2, fig.width=6, fig.align="center", fig.cap = "Multiple execution of the MH algorithm using $\\chi(\\lfloor{X_t + 1}\\rfloor)$ as the proposal distribution."}
# Create a 2x2 grid layout
par(mfrow = c(2, 2), mar = c(3, 3, 0.5, 0.5), oma = c(0, 0, 0, 0))  # Adjust margins

# Plot the first vector (x1)
plot(x1, type = "l", main = "",xlab = "Iteration",  ylab = "Value", col = "black")

# Plot the second vector (x2)
plot(x2, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Plot the third vector (x3)
plot(x3, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Plot the fourth vector (x4)
plot(x4, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Reset layout to single plot (optional)
par(mfrow = c(1, 1))
```

```{r fig.height=2, fig.width=3, fig.align="center", fig.cap="Histogram of MH deviates values using $\\chi(\\lfloor{X_t + 1}\\rfloor)$ as the proposal distribution. The red curve shows the target (true) density.", echo=FALSE}
par(mar = c(3, 3, 1, 1))  # Adjust margins: c(bottom, left, top, right)
x <- rmh(100000, densityTarget, chiDensityProposal, chiRandomProposal, 0.1)
hist(x, probability = TRUE, breaks = 20, col = "skyblue", main = "")
curve(densityTarget(x), add = TRUE, col = "red", lwd = 2)
xChi <- x
```


**What can you guess about the convergence of the chain? If there
is a burn–in period, what can be the size of this period? What is the acceptance rate?**

## Uniform distribution

```{r include=FALSE}
unDensityProposal <- function(x, y) dunif(x, y - 1, y + 1)
unRandomProposal <- function(x) runif(n = 1, x - 1, x + 1)
```

In what follows we experiment with the uniform distribution centered around 
$X_t$. In particular, $\mathcal{U}(X_t - 1, X_t + 1)$.


```{r include=FALSE}
x1 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, 0.1)
x2 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, 0.2)
x3 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, 0.3)
x4 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, 0.4)
```


```{r echo=FALSE, fig.height=2, fig.width=6, fig.align="center", fig.cap = "Multiple execution of the MH algorithm using $\\mathcal{U}(X_t - 1, X_t + 1)$ as the proposal distribution."}
# Create a 2x2 grid layout
par(mfrow = c(2, 2), mar = c(3, 3, 0.5, 0.5), oma = c(0, 0, 0, 0))  # Adjust margins

# Plot the first vector (x1)
plot(x1, type = "l", main = "",xlab = "Iteration",  ylab = "Value", col = "black")

# Plot the second vector (x2)
plot(x2, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Plot the third vector (x3)
plot(x3, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Plot the fourth vector (x4)
plot(x4, type = "l", main = "", xlab = "Iteration", ylab = "Value", col = "black")

# Reset layout to single plot (optional)
par(mfrow = c(1, 1))
```

```{r fig.height=2, fig.width=3, fig.align="center", fig.cap="Histogram of MH deviates values using $\\mathcal{U}(X_t - 1, X_t + 1)$ as the proposal distribution. The red curve shows the target (true) density.", echo=FALSE}
par(mar = c(3, 3, 1, 1))  # Adjust margins: c(bottom, left, top, right)
x <- rmh(100000, densityTarget, unDensityProposal, unRandomProposal, 0.1)
hist(x, probability = TRUE, breaks = 20, col = "skyblue", main = "")
curve(densityTarget(x), add = TRUE, col = "red", lwd = 2)
xUniform <- x
```

**What can you guess about the convergence of the chain? If there is a burn–in period, what can be the size of this period? What is the acceptance rate?**

## Comparisson


## Estimated mean

The following table shows the mean for each case.

```{r echo=FALSE}
meanDf <- data.frame(
  Normal = mean(xNormal),
  ChiSq = mean(xChi),
  Uniform = mean(xUniform)
)

knitr::kable(
  round(meanDf, 3),
  row.names = FALSE,
  caption = "Estimated mean for each of the three proposal distributions."
)
```


## True mean

We recognize $f$ to be the density of a Gamma distribution with parameters $\alpha=6$ and $\lambda=1$ whose mean is given by

$$
\mu_{\Gamma(\alpha, \lambda)} = \frac{\alpha}{\lambda} = \frac{6}{1} = 6
$$







