---
title: "Question 1: Computations with Metropolis–Hastings"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
tables: true
header-includes:
 \usepackage{float}
 \setlength{\textfloatsep}{0pt}

---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

```{r include=FALSE}
library(ggplot2)
library(gridExtra)
set.seed(2)
```


```{r include=FALSE}
densityTarget <- function(x) ifelse(x > 0, (1/120) * x^5 * exp(-x), 0)

#' Metropolis-Hastings Algorithm
#'
#' @param n integer.
#'   The number of samples to generate.
#' @param dtarget function.
#'   The target distribution's density function. 
#'   Must be a function with signature x -> dtarget(x).
#' @param dprop function.
#'   Density function for the proposed conditional distribution. 
#'   Must be a function with signature x, y -> dprop(x, y), where y is the 
#'   conditioning event and x is a value in the support.
#' @param rprop function.
#'   Random generator for the proposed conditional distribution. 
#'   Must be a function with signature x -> rprop(x), where x is the 
#'   conditioning event.
#' @param x0 numeric.
#'   Initial value for the sampling process. Must be in the support of `dprop`.
#'
#' @return list
rmh <- function(n, densityTarget, densityProposal, randomProposal, x0) {
  
  sample <- numeric(n)
  n_accepted <- 0
  new_x <- x0
   
  for (i in 1:n) {
    
    last_x <- new_x
    
    # Sample a candidate from the proposal distribution.
    y <- randomProposal(last_x)
    
    # Compute the MH ratio.
    R <- (densityTarget(y) * densityProposal(last_x, y)) / (densityTarget(last_x) * densityProposal(y, last_x))
    
    # Decide new sample.
    # Increment count if proposed sample is accepted.
    new_x <- ifelse(
      R >= 1, 
      {n_accepted <- n_accepted + 1; y},
      ifelse(runif(1) < R, {n_accepted <- n_accepted + 1; y}, last_x)
      )
    
    sample[i] <- new_x
  }
  
  acceptance_rate = n_accepted / n
  return(list(sample=sample, acceptange_rate=acceptance_rate))
}
```



We are given the target density

$$
f(x) = \frac{1}{120} x^5 e^{-x} \quad x > 0
$$

In what follows we will use the Metropolis-Hasting algorithm (MH) under
different proposal distributions to generate random samples from this 
density. In particular, the proposal distribution used are

1. $\mathcal{N}(\mu = X_t, \sigma = 0.1)$
2. $\chi(\lfloor{X_t + 1}\rfloor)$
3. $\mathcal{U}(X_t - 1, X_t + 1)$

Further details are given on each section.



## Normal distribution

```{r include=FALSE}
normalDensityProposal <- function(x, y) dnorm(x, mean = y, sd = 0.1)
normalRandomProposal <- function(x) rnorm(n = 1, mean = x, sd = 0.1)
```

Here, we apply the MH algorithm using $\mathcal{N}(\mu = X_t, \sigma = 0.1)$ as the
proposed conditional density $g_{X_{t+1} | X_t}$. That is, 

$$
g_{X_{t+1} | X_t}(x_{t+1} | x_t) = \mathcal{N}(x_{t+1} | \mu = x_t, \sigma = 0.1)
$$

The following figure shows four runs of length $10,000$ using different 
starting values. The acceptance rate is also shown at the top of each
plot.


```{r include=FALSE}
starting_values <- c(0.1, 0.2, 0.3, 0.4)
x1 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[1])
x2 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[2])
x3 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[3])
x4 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[4])

normalAcceptanceRate <- mean(x1$acceptange_rate, x2$acceptange_rate, x3$acceptange_rate, x4$acceptange_rate)
```

```{r echo=FALSE, fig.height=2, fig.width=6, fig.align="center", fig.cap = "Multiple execution of the MH algorithm using $\\mathcal{N}(\\mu = X_t, \\sigma = 0.1)$ as the proposal distribution."}
# Create a 2x2 grid layout
par(mfrow = c(2, 2), mar = c(3, 3, 0.5, 0.5), oma = c(0, 0, 0, 0), cex.main = 0.5)  # Adjust margins

# Plot the first vector (x1)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[1], x1$acceptange_rate)
plot(x1$sample, type = "l", main = main ,xlab = "Iteration",  ylab = "Value", col = "black")

# Plot the second vector (x2)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[2], x2$acceptange_rate)
plot(x2$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Plot the third vector (x3)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[3], x3$acceptange_rate)
plot(x3$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Plot the fourth vector (x4)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[4], x4$acceptange_rate)
plot(x4$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Reset layout to single plot (optional)
par(mfrow = c(1, 1))
```


## Chi-square distribution

```{r include=FALSE}
chiDensityProposal <- function(x, y) dchisq(x, df = floor(y + 1))
chiRandomProposal <- function(x) rchisq(n = 1, df = floor(x + 1))
```


Now, we use $\chi(\lfloor{X_t + 1}\rfloor)$ as the proposal distribution where
$\lfloor{\cdot}\rfloor$ is the floor function. The results are shown in the 
following figures.

```{r include=FALSE}
starting_values <- c(0.1, 0.2, 0.3, 0.4)
x1 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[1])
x2 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[2])
x3 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[3])
x4 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[4])

chiAcceptanceRate <- mean(x1$acceptange_rate, x2$acceptange_rate, x3$acceptange_rate, x4$acceptange_rate)
```

```{r echo=FALSE, fig.height=2, fig.width=6, fig.align="center", fig.cap = "Multiple execution of the MH algorithm using $\\chi(\\lfloor{X_t + 1}\\rfloor)$ as the proposal distribution."}
# Create a 2x2 grid layout
par(mfrow = c(2, 2), mar = c(3, 3, 0.5, 0.5), oma = c(0, 0, 0, 0), cex.main = 0.5)  # Adjust margins

# Plot the first vector (x1)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[1], x1$acceptange_rate)
plot(x1$sample, type = "l", main = main ,xlab = "Iteration",  ylab = "Value", col = "black")

# Plot the second vector (x2)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[2], x2$acceptange_rate)
plot(x2$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Plot the third vector (x3)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[3], x3$acceptange_rate)
plot(x3$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Plot the fourth vector (x4)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[4], x4$acceptange_rate)
plot(x4$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Reset layout to single plot (optional)
par(mfrow = c(1, 1))
```

## Uniform distribution

```{r include=FALSE}
unDensityProposal <- function(x, y) dunif(x, y - 1, y + 1)
unRandomProposal <- function(x) runif(n = 1, x - 1, x + 1)
```

Next, we experiment with the uniform distribution centered around 
$X_t$. In particular, $\mathcal{U}(X_t - 1, X_t + 1)$.


```{r include=FALSE}
starting_values <- c(0.1, 0.2, 0.3, 0.4)
x1 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[1])
x2 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[2])
x3 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[3])
x4 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[4])

uniAcceptanceRate <- mean(x1$acceptange_rate, x2$acceptange_rate, x3$acceptange_rate, x4$acceptange_rate)
```


```{r echo=FALSE, fig.height=2, fig.width=6, fig.align="center", fig.cap = "Multiple execution of the MH algorithm using $\\mathcal{U}(X_t - 1, X_t + 1)$ as the proposal distribution."}
# Create a 2x2 grid layout
par(mfrow = c(2, 2), mar = c(3, 3, 0.5, 0.5), oma = c(0, 0, 0, 0), cex.main = 0.5)  # Adjust margins

# Plot the first vector (x1)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[1], x1$acceptange_rate)
plot(x1$sample, type = "l", main = main ,xlab = "Iteration",  ylab = "Value", col = "black")

# Plot the second vector (x2)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[2], x2$acceptange_rate)
plot(x2$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Plot the third vector (x3)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[3], x3$acceptange_rate)
plot(x3$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Plot the fourth vector (x4)
main <- sprintf("Starting value=%s ; Acceptance rate=%s", starting_values[4], x4$acceptange_rate)
plot(x4$sample, type = "l", main = main, xlab = "Iteration", ylab = "Value", col = "black")

# Reset layout to single plot (optional)
par(mfrow = c(1, 1))

```


## Histograms

In addition to the sample plots from each section, we show the histogram
for one of the runs for each case along with the target (true) density.

```{r fig.height=2, fig.width=6, fig.align="center", fig.cap="Histograms for each proposal distribution. The red curve shows the target (true) density.", echo=FALSE}

n <- 10000
x <- 0:100

xNormal <- rmh(n, densityTarget, normalDensityProposal, normalRandomProposal, 1)$sample 
xChiSq <- rmh(n, densityTarget, chiDensityProposal, chiRandomProposal, 1)$sample 
xUniform <- rmh(n, densityTarget, chiDensityProposal, chiRandomProposal, 1)$sample 


# Create the histograms
par(mfrow = c(1, 3))
hist(xNormal, probability = TRUE, breaks = 20, col = "skyblue", main = "")
curve(densityTarget(x), add = TRUE, col = "red", lwd = 2)

hist(xChiSq, probability = TRUE, breaks = 20, col = "skyblue", main = "")
curve(densityTarget(x), add = TRUE, col = "red", lwd = 2)

hist(xUniform, probability = TRUE, breaks = 20, col = "skyblue", main = "")
curve(densityTarget(x), add = TRUE, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```


## Comparisson: convergence, burn-in period and acceptance rate

The following table summarizes the distinctions between the three
proposal distribution.

```{r echo=FALSE}

acceptance_rates <- c(normalAcceptanceRate, chiAcceptanceRate, uniAcceptanceRate)
burn_in_period <- c("0-1000", "0-10", "0-100")
comparissonDf <- data.frame(
  acceptanceRate = acceptance_rates,
  burnInPeriod = burn_in_period,
  convergence = c("Yes", "Yes", "Yes")
)

rownames(comparissonDf) <- c("$N$", "$\\chi$", "$U$")


knitr::kable(
  t(comparissonDf),
  caption = "Comparisson between the proposal distributions"
)


```

The acceptance rate what calculated using the mean of each of the $4$ runs
for each distribution. Both $\mathcal{N}$ and $\mathcal{U}$ show high 
acceptance range which is often not desirable since it means the algorithm
is not being exploratory enough (too conservative). According to the histograms, all distributions seem to converge to the target (true) density; however, the trace plot from$\mathcal{N}$ indicates the steps taken are too small due to the low standard deviation (which also explain the high acceptance rate), Hence, convergence for this case takes longer. Burn-in periods were determined visually, by the trace plots. Here the $\chi$ distribution shows little to no burn-in period, meaning convergence is faster than the rest.
 



## Estimated mean

The following table shows the mean for each case.

```{r echo=FALSE}

meanDf <- data.frame(
  Normal = mean(xNormal),
  ChiSq = mean(xChiSq),
  Uniform = mean(xUniform)
)

names(meanDf) <- c("$N$", "$\\chi$", "$U$")

knitr::kable(
  round(meanDf, 3),
  row.names = FALSE,
  caption = "Estimated mean for each of the three proposal distributions."
)
```


## True mean

We recognize $f$ to be the density of a Gamma distribution with parameters $\alpha=6$ and $\lambda=1$ whose mean is given by

$$
\mu_{\Gamma(\alpha, \lambda)} = \frac{\alpha}{\lambda} = \frac{6}{1} = 6
$$


## Appendix


```r

densityTarget <- function(x) ifelse(x > 0, (1/120) * x^5 * exp(-x), 0)

#' Metropolis-Hastings Algorithm
#'
#' @param n integer.
#'   The number of samples to generate.
#' @param dtarget function.
#'   The target distribution's density function. 
#'   Must be a function with signature x -> dtarget(x).
#' @param dprop function.
#'   Density function for the proposed conditional distribution. 
#'   Must be a function with signature x, y -> dprop(x, y), where y is the 
#'   conditioning event and x is a value in the support.
#' @param rprop function.
#'   Random generator for the proposed conditional distribution. 
#'   Must be a function with signature x -> rprop(x), where x is the 
#'   conditioning event.
#' @param x0 numeric.
#'   Initial value for the sampling process. Must be in the support of `dprop`.
#'
#' @return list
rmh <- function(n, densityTarget, densityProposal, randomProposal, x0) {
  
  sample <- numeric(n)
  n_accepted <- 0
  new_x <- x0
   
  for (i in 1:n) {
    
    last_x <- new_x
    
    # Sample a candidate from the proposal distribution.
    y <- randomProposal(last_x)
    
    # Compute the MH ratio.
    R <- (densityTarget(y) * densityProposal(last_x, y)) / (densityTarget(last_x) * densityProposal(y, last_x))
    
    # Decide new sample.
    # Increment count if proposed sample is accepted.
    new_x <- ifelse(
      R >= 1, 
      {n_accepted <- n_accepted + 1; y},
      ifelse(runif(1) < R, {n_accepted <- n_accepted + 1; y}, last_x)
      )
    
    sample[i] <- new_x
  }
  
  acceptance_rate = n_accepted / n
  return(list(sample=sample, acceptange_rate=acceptance_rate))
}


#  ---- NORMAL DISTRIBUTION ----

normalDensityProposal <- function(x, y) dnorm(x, mean = y, sd = 0.1)
normalRandomProposal <- function(x) rnorm(n = 1, mean = x, sd = 0.1)
starting_values <- c(0.1, 0.2, 0.3, 0.4)
x1 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[1])
x2 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[2])
x3 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[3])
x4 <- rmh(10000, densityTarget, normalDensityProposal, normalRandomProposal, starting_values[4])
normalAcceptanceRate <- mean(x1$acceptange_rate, x2$acceptange_rate, x3$acceptange_rate, x4$acceptange_rate)


#  ---- CHISQ DISTRIBUTION ----

chiDensityProposal <- function(x, y) dchisq(x, df = floor(y + 1))
chiRandomProposal <- function(x) rchisq(n = 1, df = floor(x + 1))
starting_values <- c(0.1, 0.2, 0.3, 0.4)
x1 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[1])
x2 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[2])
x3 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[3])
x4 <- rmh(10000, densityTarget, chiDensityProposal, chiRandomProposal, starting_values[4])
chiAcceptanceRate <- mean(x1$acceptange_rate, x2$acceptange_rate, x3$acceptange_rate, x4$acceptange_rate)

# ---- UNIFORM DISTRIBUTION ----

unDensityProposal <- function(x, y) dunif(x, y - 1, y + 1)
unRandomProposal <- function(x) runif(n = 1, x - 1, x + 1)
starting_values <- c(0.1, 0.2, 0.3, 0.4)
x1 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[1])
x2 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[2])
x3 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[3])
x4 <- rmh(10000, densityTarget, unDensityProposal, unRandomProposal, starting_values[4])
uniAcceptanceRate <- mean(x1$acceptange_rate, x2$acceptange_rate, x3$acceptange_rate, x4$acceptange_rate)
```


