---
title: "Question 2: Simulation of power curves"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
header-includes:
 \usepackage{float}
 \setlength{\textfloatsep}{0pt}
 \usepackage{hyperref}
tables: true
colorlinks: true
linkcolor: red
---

```{r include=FALSE}
library(BSDA)
```


```{r include=FALSE}
pgumbel <- function(q, loc = 0, scale = 1) exp(-exp(-(x - loc) / scale))
qgumbel <- function(p, loc = 0, scale = 1) loc - scale * log(-log(p))
rgumbel <- function(n, loc = 0, scale = 1) qgumbel(runif(n), loc, scale)
dgumbel <- function(x, loc = 0, scale = 1) {
  z <- (x - loc) / scale
  (1 / scale) * exp(-(z + exp(-z)))
}
```

The Gumbel distribution has the distribution function

\begin{equation}
\label{eq:Gumbel}
F(x ; \mu, \beta) = \exp{\left\{-\exp{\left\{\frac{-(x - \mu)}{\beta}\right\}}\right\}} 
\end{equation}


where the parameters $\mu$ and $\beta > 0$ represent the location and scale, respectively. The median $M$ of this distribution is $\mu - \beta \log(\log(2))$.


## Sampling using the Inverse Transform Method

Let $X \sim F$. Then, we can draw from $X$ through the transformation 
$X = F^{-1}(U)$ where $U \sim \mathcal{U}(0, 1)$ and $F^{-1}$ is the inverse
of $F$. For the Gumbel distribution, $F$ is given by \ref{eq:Gumbel} and its inverse
is obtained by solving for $x$:


$$
\begin{aligned}
 F(x)                                      &= \exp{\left\{-\exp{\left\{\frac{-(x - \mu)}{\beta}\right\}}\right\}}  \\
-\log{F(x)}                                &= \exp{\left\{\frac{-(x - \mu)}{\beta}\right\}} \\
-\beta\log\left\{-\log{F(x)}\right\} + \mu &= x
\end{aligned}
$$

Thus, $X = F^{-1}(U) = -\beta\log\left\{-\log{U}\right\} + \mu$. The following histogram
shows the distribution of $X$ using this transformation for $\mu=\log(\log(2))$ and $\beta=1$ (i.e., median is zero) along with the true density function:

```{r fig.height=2, fig.width=2.5, fig.align="center", fig.cap="Histogram of Gumbel draws obtained using the inverse transform method for $\\mu=\\log(\\log(2))$ and $\\beta=1$.", echo=FALSE}
par(mar = c(3, 3, 1, 1))  # Adjust margins: c(bottom, left, top, right)
x = rgumbel(1000, log(log(2)))
hist(x, probability = TRUE, breaks = 10, col = "gray", main = "")
curve(dgumbel(x), add = TRUE, col = "red", lwd = 2)
```


## Power of the Sign test

Let $X_1, \dots, X_n$ be a random sample of size $n$ with median $M$. We aim to investigate the power (i.e., probability of rejecting $H_0$) of the Sign test for:

$$
H_0: M = 0 \quad \text{versus} \quad H_a: M > 0
$$


To do so, rather than deriving power analytically, we estimate the power by drawing $n$ samples from any proposal distribution with $M > 0$ (i.e., under $H_a$) and computing the rejection rate over $s$ repetitions. To generate the power curve, this procedure is repeated for multiple values of $M$. The following plot shows the
results from this experiment using the Gumbel distribution with $\mu = M + \log(\log2)$ and $\beta=1$ as the proposal.




```{r include=FALSE}

M <- seq(from = 0, to = 2, by = 0.1)
P = numeric(length(M))
s <- 1000
n <- 13
c <- log(log(2))

for (i in 1:length(M)) {
  m <- M[i]
  count <- 0
  
  for (sim in 1:s) {
    
    # True distribution has M > 0
    loc = m + c
    x <- rgumbel(n, loc = loc)
    
    # Test with M = 0
    reject <- (SIGN.test(x, alternative = "greater")$p.value < 0.05)
    count <- count + reject
  }
  
  P[i] <- count / s
}

```

```{r, fig.height=2, fig.width=4, fig.align="center", fig.cap="Power curve of the sign test using Gumbell as proposal.",  echo=FALSE}
par(mar = c(3, 3, 1, 1))  # Adjust margins: c(bottom, left, top, right)
plot(M, P, type = "o", col = "blue", xlab = "Median (M)", ylab = "Power")
```

As expected, the Type 2 error (probability of accepting $H_0$ given $H_0$ is false) decreases as $M$ moves further from 0. In other words, the evidence *against* $M = 0$ strengthens as $M$ increases and, by $M = 0.7$, we reject $H_0$ more often than not.

## Appendix

```r
library(BSDA)

# --- GUMBEL ---
pgumbel <- function(q, loc = 0, scale = 1) exp(-exp(-(x - loc) / scale))
qgumbel <- function(p, loc = 0, scale = 1) loc - scale * log(-log(p))
rgumbel <- function(n, loc = 0, scale = 1) qgumbel(runif(n), loc, scale)
dgumbel <- function(x, loc = 0, scale = 1) {
  z <- (x - loc) / scale
  (1 / scale) * exp(-(z + exp(-z)))
}

# --- POWER CURVE ---
M <- seq(from = 0, to = 2, by = 0.1)
P = numeric(length(M))
s <- 1000
n <- 13
c <- log(log(2))

for (i in 1:length(M)) {
  m <- M[i]
  count <- 0
  
  for (sim in 1:s) {
    
    # True distribution has M > 0
    loc = m + c
    x <- rgumbel(n, loc = loc)
    
    # Test with M = 0
    reject <- (SIGN.test(x, alternative = "greater")$p.value < 0.05)
    count <- count + reject
  }
  
  P[i] <- count / s
}
```




