---
title: "Question 2: Bivariate normal and normal mixture distribution"
output:
  pdf_document: 
    keep_tex: true
    latex_engine: xelatex
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
tables: true
header-includes:
 \usepackage{float}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
library(mvtnorm)
```

```{r include=FALSE}
#' Box_Muller method for generating norm distribution
#'
#' @param n Number of samples
#' @param mu Mean vector of distribution
#' @param sigma Covariance matrix of distribution
#'
#' @returns Matrix of samples
box_muller <- function(n,mu,sigma){
  u <- runif(n,0,1)
  A <- runif(n,0,2*pi)
  x1 <- sqrt(-2*log(u)) * cos(A)
  x2 <- sqrt(-2*log(u)) * sin(A)
  x_standard <- cbind(x1,x2)
  w <- chol(sigma)
  Y <- x_standard%*%t(w) + matrix(rep(mu, each = n), ncol = 2)
  return(Y)
}

#' Normal mixture distribution generator 
#'
#' @param n Number of samples
#' @param prob The probability that sample belongs to the first distribution
#' @param mu1 Mean vector of first distribution
#' @param mu2 Mean vector of second distribution
#' @param sigma1 Covariance matrix of first distribution
#' @param sigma2 Covariance matrix of second distribution
#'
#' @returns Matrix of samples
mix_normal <- function(n,prob,mu1,mu2,sigma1,sigma2){
  
  # get the number of observations of each density
  density1_n <- floor(n*prob)
  density2_n <- n-density1_n
  
  sample1 <- box_muller(density1_n,mu1,sigma1)
  sample2 <- box_muller(density2_n,mu2,sigma2)
  
  # add lable to each component
  sample1 <- cbind(sample1,1)
  sample2 <- cbind(sample2,2)
  
  samples <- rbind(sample1,sample2)
  return(samples)
}
```

The time to generate 10000000 random vectors using the box_muller method
is:

```{r echo=FALSE}
mu <- c(0,0)
sigma <- matrix(c(0.6,0,0,0.6),nrow = 2)
system.time(box_muller(10000000,mu,sigma))
```

Now we generate again 10000000 random vectors with package `mvtnorm`, because `rnorm()` can only generate one-dimensional normal distribution data. `rmvnorm()` directly generates data that conforms to the target distribution, avoiding additional matrix operations.

So the time to generate data using `rmvnorm()` is:

```{r echo=FALSE}
system.time(rmvnorm(10000000,mu,sigma,method = "chol"))
```
By comparing the results with a, we can see that using rmvnorm() is faster to generate random vectors.


Next, we generate a bivariate normal mixture distribution using two probability densities, and the plot is shown below:

```{r echo=FALSE}
mu1 <- c(0,0)
mu2 <- c(1.5,1.2)
sigma1 <- matrix(c(0.6,0,0,0.6),nrow = 2)
sigma2 <- matrix(c(0.5,0,0,0.5),nrow = 2)

new_samples <- mix_normal(1000,0.5,mu1,mu2,sigma1,sigma2)
sample1 <- new_samples[which(new_samples[,3] == 1), ]
sample2 <- new_samples[which(new_samples[,3] == 2), ]
plot(sample1[,1],sample1[,2],col="red",ylim = c(-3,3),xlim = c(-2.5,3))
points(sample2[,1],sample2[,2],col="blue")
```

The plot looks satisfactory, because the plot features satisfy the following three points:

1. The plot clearly shows two clusters with centers close to the theoretical mean(0,0) and (1.5,1.2).

2. The sample points of cluster 1 are more dispersed than cluster 2, which is consistent with the covariance difference.(0.6 vs. 0.5).

3. There are no obvious outliers or error distributions in the sample points.

## Appendix
```{r echo=TRUE, eval=FALSE}
#' Box_Muller method for generating norm distribution
#'
#' @param n Number of samples
#' @param mu Mean vector of distribution
#' @param sigma Covariance matrix of distribution
#'
#' @returns Matrix of samples
box_muller <- function(n,mu,sigma){
  u <- runif(n,0,1)
  A <- runif(n,0,2*pi)
  x1 <- sqrt(-2*log(u)) * cos(A)
  x2 <- sqrt(-2*log(u)) * sin(A)
  x_standard <- cbind(x1,x2)
  w <- chol(sigma)
  Y <- x_standard%*%t(w) + matrix(rep(mu, each = n), ncol = 2)
  return(Y)
}

#' Normal mixture distribution generator 
#'
#' @param n Number of samples
#' @param prob The probability that sample belongs to the first distribution
#' @param mu1 Mean vector of first distribution
#' @param mu2 Mean vector of second distribution
#' @param sigma1 Covariance matrix of first distribution
#' @param sigma2 Covariance matrix of second distribution
#'
#' @returns Matrix of samples
mix_normal <- function(n,prob,mu1,mu2,sigma1,sigma2){
  
  # get the number of observations of each density
  density1_n <- floor(n*prob)
  density2_n <- n-density1_n
  
  sample1 <- box_muller(density1_n,mu1,sigma1)
  sample2 <- box_muller(density2_n,mu2,sigma2)
  
  # add lable to each component
  sample1 <- cbind(sample1,1)
  sample2 <- cbind(sample2,2)
  
  samples <- rbind(sample1,sample2)
  return(samples)
}

mu <- c(0,0)
sigma <- matrix(c(0.6,0,0,0.6),nrow = 2)

system.time(box_muller(10000000,mu,sigma))
system.time(rmvnorm(10000000,mu,sigma,method = "chol"))

mu1 <- c(0,0)
mu2 <- c(1.5,1.2)
sigma1 <- matrix(c(0.6,0,0,0.6),nrow = 2)
sigma2 <- matrix(c(0.5,0,0,0.5),nrow = 2)

new_samples <- mix_normal(1000,0.5,mu1,mu2,sigma1,sigma2)
sample1 <- new_samples[which(new_samples[,3] == 1), ]
sample2 <- new_samples[which(new_samples[,3] == 2), ]
plot(sample1[,1],sample1[,2],col="red",ylim = c(-3,3),xlim = c(-3,3))
points(sample2[,1],sample2[,2],col="blue")
```

